\textbf{Predicting virus-receptor mutant binding by molecular dynamics simulation}

### 2013-07-09 paper.tex
Updated everything to current text.
Changes to manuscript.

\begin{abstract}

Existing computational methods to predict protein--protein interaction affinity often perform poorly in important test cases. In particular, the effects of multiple mutations, non-alanine substitutions, and flexible loops are difficult to predict with available tools and protocols. We present here a new method to interrogate affinity differences resulting from mutations in a host-virus protein--protein interface. Our method is based on extensive non-equilibrium all-atom simulations: We computationally pull the machupo virus (MACV) spike glycoprotein (GP1) away from the human transferrin receptor (hTfR1) and estimate affinity using the maximum applied force during a pulling simulation and the area under the force-versus-distance curve. We find that these quantities can provide novel biophysical insight into the GP1/hTfR1 interaction. First, with no prior knowledge of the system we can differentiate among wild-type and mutant complexes. Second, although the static co-crystal structure shows two large hydrogen-bonding networks in the GP1/hTfR1 interface, our simulations indicate that one of them may not be important for tight binding. Third, one viral site known to be critical for infection may mark an important evolutionary suppressor site for infection-resistant hTfR1 mutants. Finally, our method provides an elegant framework to compare the effects of multiple mutations, individually and jointly, on protein--protein interactions.
\end{abstract}

@@

The computational prediction of mutational effects on protein--protein interactions remains a challenging problem. Several methods are available to perform an energy difference calculation from an experimentally determined co-crystal structure. For example, computational alanine-scanning can be performed rapidly, with relatively low computational cost \citep{Grant2011,Kortemme2004}. However, such methods calculate the relative contribution to overall energy for each amino acid in a protein. They do not probe the interface directly, but rather calculate the total energy with each substitution \citep{Grant2011,Kortemme2004}. Alternative approaches have been developed using machine learning, training coefficients in a weighted equation containing geometric and energetic parameters \citep{Vreven2011,Vreven2012,Bajaj2011,Hwang2010}. Unfortunately, such machine-learning approaches often suffer in novel applications, for which available training sets are small or non-existent. As such, these methods are poorly suited for most host-virus protein--protein systems. By contrast, first principles methods offer the chance to forgo training. However, the currently available methods such as free energy perturbation (FEP) and thermodynamic integration (TI) rely on a two state model (where one state may be wild-type and the other may be a mutant) that transitions directly from one three-dimensional protein model to another with no intermediate steps \citep{Gilson1997,Lu2004}. As a result, the accuracy of the measurement suffers as the two proteins become increasingly dissimilar; therefore, these techniques are largely limited to point mutant comparisons.

Here we propose a new method to probe mutational effects in protein--protein interactions directly based on first principles simulation, by pulling two proteins apart that are initially bound in a complex. We impart a pulling force within an all-atom molecular dynamics simulation on one member of the complex while the other is held in place. Then, we measure the force required for dissociation. As a proxy for relative binding affinity, we use the maximum applied force required for separation and the area under the force-versus-distance curve (AUC).

@@

In every way possible the subsequent steering simulation had the same parameters as the equilibration simulation. Thus, the steered molecular dynamics (SMD) \citep{Is2001B} simulation was a typical restart with slightly different settings. Periodic boundary conditions were incorporated as part of the system and PME was again used to approximate full system electrostatics. In addition, we first tuned the pulling velocity for practical time considerations. Subsequently, the spring constant was adjusted to ensure a sufficient signal-to-noise ratio with the given velocity. As a result, the spring constant was set to 5~kcal/mol/\AA$^2$, velocity set to 1~\AA/ns, and direction down the positive $z$-axis.

@@

\subsection*{Comparative analysis of the GP1/hTfR1 interface}
Considering the involvement of extended hydrogen-bonding networks in the GP1/hTfR1 interface, it was not clear that individual alanine mutations, even those that should destroy such networks, would significantly change the strength of interaction. One major advantage of first principles simulations is the ability to test mutations other than alanine without additional underlying assumptions in the energy function. Seen in Table~\ref{tab:summary_mutant}), we made additional mutations based on biochemical intuition or available experimental data to chemically diverse amino acids including tryptophan, lysine, aspartate, and threonine. Several mutations caused significant relative affinity changes. In addition, to detect synergistic effects, we tested several double mutants where both mutations appeared to cause similar changes in binding. Then, we compared the size of those differences to single mutants (Figure~\ref{fig:sensitivity} and ~\ref{fig:mutant_comparison}).

@@

Last, we further analyzed a single mutation in GP1, vArg111Ala. As mentioned in the previous subsection, in our simulations this mutant showed no significant change in either maximum applied force or AUC (Tables~\ref{tab:MAF_p} and~\ref{tab:AUC_p}), even though both quantities were, on average, lower than in WT (Table~\ref{tab:summary_statistics}). This result was somewhat surprising, since Tyr211Ala, presumably disrupting the same hydrogen-bonding network as vArg111Ala, displayed a significant reduction in affinity. To probe the interaction between position 111 in the GP1 and position 211 in the hTfR1 further, we also tested the double mutant vArg111Ala/Tyr211Ala. This double mutant showed affinity indistinguishable from WT and significantly higher than Tyr211Ala alone  (Table~\ref{tab:MAF_pairwises}). This result shows that the two sites do indeed interact, and that replacing the hydrogen-bonding network at these sites with a hydrophobic interaction could lead to comparable binding affinity.

@@

This work was supported by the Defense Threat Reduction Agency (HDTRA1-12-C-0007) to A.D.E., S.L.S., and C.O.W., the National Science Foundation (MCB-0943383) to A.D.E., the National Institutes of Health (R01-GM088344) to C.O.W., and the National Institutes of Health (R01-GM093086) to S.L.S.

>>>>

\begin{abstract}
Existing computational methods to predict protein--protein interaction affinity often perform poorly in important test cases. In particular, the effects of multiple mutations, non-alanine substitutions, and flexible loops are difficult to predict with available tools and protocols. We present here a new method to interrogate affinity differences resulting from mutations in a host-virus protein--protein interface. Our method is based on extensive non-equilibrium all-atom simulations: We computationally pull the machupo virus (MACV) spike glycoprotein (GP1) away from the human transferrin receptor (hTfR1) and estimate affinity using the maximum applied force during a pulling simulation and the area under the force-versus-distance curve. We find that these quantities can provide novel biophysical insight into the GP1/hTfR1 interaction. First, with no prior knowledge of the system we can differentiate among wild-type and mutant complexes. Second, although the static co-crystal structure shows two large hydrogen-bonding networks in the GP1/hTfR1 interface, our simulations indicate that one of them may not be important for tight binding. Third, one viral site known to be critical for infection may mark an important evolutionary suppressor site for infection-resistant hTfR1 mutants. Finally, our method provides an elegant framework to compare the effects of multiple mutations, individually and jointly, on protein--protein interactions.
\end{abstract}

@@

The computational prediction of mutational effects on protein--protein interactions remains a challenging problem. Several methods are available to perform an energy difference calculation from an experimentally determined co-crystal structure. For example, computational alanine-scanning can be performed rapidly, with relatively low computational cost \citep{Grant2011,Kortemme2004}. However, such methods calculate the relative contribution to overall energy for each amino acid in a protein. They do not probe the interface directly, but rather calculate the total energy with each substitution \citep{Grant2011,Kortemme2004}. Alternative approaches have been developed using machine learning, training coefficients in a weighted equation containing geometric and energetic parameters \citep{Vreven2011,Vreven2012,Bajaj2011,Hwang2010}. Unfortunately, such machine-learning approaches often suffer in novel applications, for which available training sets are small or non-existent. As such, these methods are poorly suited for most host-virus protein--protein systems. By contrast, first principles methods offer the chance to forgo training. However, the currently available methods such as free energy perturbation (FEP) and thermodynamic integration (TI) rely on a two state model (where one state may be wild-type and the other may be a mutant) that transitions directly from one three-dimensional model to another with no intermediate steps \citep{Gilson1997,Lu2004,Chodera2011,Gumbart2013}. As a result, the accuracy of the measurement suffers as the two proteins become increasingly dissimilar; therefore, these techniques are largely limited to point mutant comparisons.

Here we propose a new method to probe mutational effects in protein--protein interactions directly based on first principles simulation, by pulling two proteins apart that are initially bound in a complex. We impart a pulling force within an all-atom molecular dynamics simulation on one member of the complex while the other is held in place. Then, we measure the force required for dissociation \citep{Gumbart2012,Lu1999,Park2004,Is2001A,Is2001B}. As a proxy for relative binding affinity, we use the maximum applied force required for separation and the area under the force-versus-distance curve (AUC).

@@

In every way possible the subsequent steering simulation had the same parameters as the equilibration simulation. Thus, the steered molecular dynamics (SMD) \citep{Is2001A,Is2001B} simulation was a typical restart with slightly different settings. Periodic boundary conditions were incorporated as part of the system and PME was again used to approximate full system electrostatics. In addition, we first tuned the pulling velocity for practical time considerations. Subsequently, the spring constant was adjusted to ensure a sufficient signal-to-noise ratio with the given velocity. As a result, the spring constant was set to 5~kcal/mol/\AA$^2$, velocity set to 1~\AA/ns, and direction down the positive $z$-axis.

@@

\subsection*{Comparative analysis of the GP1/hTfR1 interface}
Considering the involvement of extended hydrogen-bonding networks in the GP1/hTfR1 interface, it was not clear that individual alanine mutations, even those that should destroy such networks, would significantly change the strength of interaction. One major advantage of first principles simulations is the ability to test mutations other than alanine without additional underlying assumptions in the energy function. Seen in Table~\ref{tab:summary_mutants}), we made additional mutations based on biochemical intuition or available experimental data to chemically diverse amino acids including tryptophan, lysine, aspartate, and threonine. Several mutations caused significant relative affinity changes. In addition, to detect synergistic effects, we tested several double mutants where both mutations appeared to cause similar changes in binding. Then, we compared the size of those differences to single mutants (Figure~\ref{fig:sensitivity} and ~\ref{fig:mutant_comparison}).

@@

Last, we further analyzed a single mutation in GP1, vArg111Ala. As mentioned in the previous subsection, in our simulations this mutant showed no significant change in either maximum applied force or AUC (Tables~\ref{tab:MAF_p} and~\ref{tab:AUC_p}), even though both quantities were, on average, lower than in WT (Table~\ref{tab:summary_statistics}). This result was somewhat surprising, since Tyr211Ala, presumably disrupting the same hydrogen-bonding network as vArg111Ala, displayed a significant reduction in affinity. To probe the interaction between position 111 in the GP1 and position 211 in the hTfR1 further, we also tested the double mutant vArg111Ala/Tyr211Ala. This double mutant showed affinity indistinguishable from WT and significantly higher than Tyr211Ala alone  (Table~\ref{tab:MAF_pairwise}). This result shows that the two sites do indeed interact, and that replacing the hydrogen-bonding network at these sites with a hydrophobic interaction could lead to comparable binding affinity.

@@

This work was supported by the Defense Threat Reduction Agency (HDTRA1-12-C-0007) to A.D.E., S.L.S., and C.O.W., the National Science Foundation (MCB-0943383) and the Welch Foundation (F-1654) to A.D.E., the National Institutes of Health (R01-GM088344) to C.O.W., and the National Institutes of Health (R01-GM093086) to S.L.S.
